---
title: "Interactive Coding Exercise - Visualisation"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  In this tutorial, we will walk you through basic plotting exercises using ggplot2 in R.
---

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
library(dplyr)
library(gradethis)

# Load dataset we wish to use here.
brfss <- read.csv("Data/dataset-brfss-2013-subset1.csv")

tutorial_options(exercise.checker = gradethis::grade_learnr)

knitr::opts_chunk$set(error = TRUE)

```

## Live Practice{#p}

Below you will find a "practice box" mimicking an R script just like one you can create in RStudio.

### Step 1: Install and load required libraries

#### Let's install and load the `ggplot2` package.

```{r load-package, exercise = TRUE}
# We have installed the package for you, now please load the library.
# Hint: Use library() to load the package.
```

```{r load-package-hint}
# Use the install.packages() function first
```

```{r load-package-solution}
# Solution;

# The load the package to utilise.
library(ggplot2)
```

```{r load-package-check}
"Great job! The package is now loaded, it is ready to use now."
```

---

### Step 2: Load the BRFSS Dataset

#### Let's load the BRFSS dataset and assign it to a variable called `brfss`.

```{r load-data, exercise = TRUE}
# Load the dataset provided and assign it to a variable called `brfss`.
# Hint: Use read.csv() to load the data.
```

```{r load-data-hint}
# The dataset is in CSV format. Use `read.csv("path/to/file.csv")` to load it.
```

```{r load-data-solution}
# Solution;
brfss <- read.csv("Data/dataset-brfss-2013-subset1.csv") # This will usually change depending on where you have your data or the format. 
```

```{r load-data-check}
"Great job! The dataset is loaded and ready for exploration."
```

---

### Step 3: Bar Plot of Education Levels

#### Create a bar plot to show the distribution of education levels in the BRFSS dataset.

```{r education-barplot, exercise = TRUE}
# Now that you have loaded the dataset and named it brfss, use ggplot2 to create a bar plot showing the distribution of the 'education' column.


```

```{r education-barplot-hint}
# Hint: ggplot('your data here', aes(x = factor('education variable here'))) + geom_bar()

# Use `geom_bar()` to create a bar plot and set the x aesthetic to `factor(education)` to ensure it is treated as categorical.
```

```{r education-barplot-solution}
# Solution
ggplot(brfss, aes(x = factor(education))) + 
  geom_bar() 
```

```{r education-barplot-check}
grade_this({
  # Check if .result is a ggplot object
  if (inherits(.result, "ggplot") &&
      # Check if the data used in the plot is `brfss`
      identical(.result$data, brfss) &&
      # Check if x and y mappings are correct
      rlang::quo_text(.result$mapping$x) == "factor(education)" &&
      # Check if there is a geom_boxplot layer
      any(sapply(.result$layers, function(layer) inherits(layer$geom, "GeomBar")))
  ) {
    pass("Well done! You've correctly plotted the bar plot for education level.")
  } else {
    fail("Try again. Ensure you use `ggplot(brfss, aes(x = factor(education))) + 
  geom_bar()`.")
  }
})
```
---

### Step 4: Histogram of BMI

#### Create a histogram to display the distribution of BMI among respondents.

```{r bmi-histogram, exercise = TRUE}
# Use ggplot2 to create a histogram of the 'bmi' column.


```

```{r bmi-histogram-hint}
# Hint: ggplot('your data here', aes(x = 'bmi variable here')) + geom_histogram(bins = 15)

# Use `geom_histogram(bins = 15)` to specify the number of bins in the histogram.
```

```{r bmi-histogram-solution}
# Solution
ggplot(brfss, aes(x = bmi)) + 
  geom_histogram(bins = 15) 
```

```{r bmi-histogram-check}
grade_this({
  # Check if .result is a ggplot object
  if (inherits(.result, "ggplot") &&
      # Check if the data used in the plot is `brfss`
      identical(.result$data, brfss) &&
      # Check if x and y mappings are correct
      rlang::quo_text(.result$mapping$x) == "bmi" &&
      # Check if there is a geom_boxplot layer
      any(sapply(.result$layers, function(layer) inherits(layer$geom, "GeomBar")))
  ) {
    pass("Well done! You've correctly plotted a histogram for BMI.")
  } else {
    fail("Try again. Ensure you use `ggplot(brfss, aes(x = bmi)) + 
  geom_histogram(bins = 15)`.")
  }
})
```
---

### Step 5: Bar Plot of Fruit and Vegetable Consumption

#### Create a bar plot to show the number of respondents consuming fruits and vegetables.

```{r diet-barplot, exercise = TRUE}
# Use ggplot2 to create a bar plot for fruit and vegetable consumption.

```

```{r diet-barplot-hint}
# Hint: ggplot('your data here', aes(x = factor('fruits variable here'))) + geom_bar()

# Set the x aesthetic to `factor(fruits)` for fruit consumption and use `geom_bar()`.
```

```{r diet-barplot-solution}
# Solution
ggplot(brfss, aes(x = factor(fruits))) + 
  geom_bar()
```


```{r diet-barplot-check}
grade_this({
  # Check if .result is a ggplot object
  if (inherits(.result, "ggplot") &&
      # Check if the data used in the plot is `brfss`
      identical(.result$data, brfss) &&
      # Check if x and y mappings are correct
      rlang::quo_text(.result$mapping$x) == "factor(fruits)" &&
      # Check if there is a geom_boxplot layer
      any(sapply(.result$layers, function(layer) inherits(layer$geom, "GeomBar")))
  ) {
    pass("Well done! You've correctly plotted the barplot for fruit and vegetable consumption.")
  } else {
    fail("Try again. Ensure you use `ggplot(brfss, aes(x = factor(fruits))) + 
  geom_bar()`.")
  }
})
```
---

### Step 6: Scatter Plot of Physical Activity and BMI

#### Create a scatter plot to explore the relationship between physical activity and BMI.

```{r activity-bmi-scatter, exercise = TRUE}
# Use ggplot2 to create a scatter plot with 'activetimes' on the x-axis and 'bmi' on the y-axis.

```

```{r activity-bmi-scatter-hint}
# Hint: ggplot('your data here', aes(x = 'activetimes variable here', y = 'bmi variable here')) + geom_point()

# Use `geom_point()` to create a scatter plot, setting 'activetimes' on the x-axis and 'bmi' on the y-axis.
```

```{r activity-bmi-scatter-solution}
# Solution
ggplot(brfss, aes(x = activetimes, y = bmi)) + 
  geom_point() 
```

```{r activity-bmi-scatter-check}
grade_this({
  # Check if .result is a ggplot object
  if (inherits(.result, "ggplot") &&
      # Check if the data used in the plot is `brfss`
      identical(.result$data, brfss) &&
      # Check if x and y mappings are correct
      rlang::quo_text(.result$mapping$x) == "activetimes" &&
      rlang::quo_text(.result$mapping$y) == "bmi" &&
      # Check if there is a geom_boxplot layer
      any(sapply(.result$layers, function(layer) inherits(layer$geom, "GeomPoint")))
  ) {
    pass("Well done! You've correctly plotted the scatterplot of activetimes and BMI.")
  } else {
    fail("Try again. Ensure you use `ggplot(brfss, aes(x = activetimes, y = bmi)) + 
  geom_point()`.")
  }
})
```
---

### Step 7: Box Plot of BMI by Education Level

#### Create a box plot to compare BMI across different education levels.

```{r bmi-education-boxplot, exercise = TRUE}
# Use ggplot2 to create a box plot of 'bmi' by 'education' levels.

```

```{r bmi-education-boxplot-hint}
# Hint: ggplot('your data here', aes(x = factor('education variable here'), y = 'bmi variable here')) + geom_boxplot() 

# Use `geom_boxplot()` to show the spread of 'bmi' across different 'education' levels.
```

```{r bmi-education-boxplot-solution}
# Solution
ggplot(brfss, aes(x = factor(education), y = bmi)) + 
  geom_boxplot() 
```

```{r bmi-education-boxplot-check}
grade_this({
  # Check if .result is a ggplot object
  if (inherits(.result, "ggplot") &&
      # Check if the data used in the plot is `brfss`
      identical(.result$data, brfss) &&
      # Check if x and y mappings are correct
      rlang::quo_text(.result$mapping$x) == "factor(education)" &&
      rlang::quo_text(.result$mapping$y) == "bmi" &&
      # Check if there is a geom_boxplot layer
      any(sapply(.result$layers, function(layer) inherits(layer$geom, "GeomBoxplot")))
  ) {
    pass("Well done! You've correctly plotted the boxplot of BMI by education level.")
  } else {
    fail("Try again. Ensure you use `ggplot(brfss, aes(x = factor(education), y = bmi)) + geom_boxplot()`.")
  }
})
```
---

This tutorial provides hands-on exercises in basic plotting with ggplot2 using the BRFSS dataset.